---
title: "Section 3: Working with Harmonised Datasets"
output:
  word_document:
    toc: true
    toc_depth: 4
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: true
    toc_depth: 6
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r session-info, include=FALSE}
## For replicability: session information 
sessionInfo()
```

```{r setup, include=FALSE}
Sys.setenv(LANG = "en")


# Define which packages needed for analyses
p_needed <-
  c(
    "knitr",
    "haven",
    "labelled",
    "dplyr",
    "tidyr",
    "gt",
    "ggplot2" ,
    "janitor",
     "here" 
    )

# Check which packages are already installed on  computer
packages <- rownames(installed.packages())


# Check which packages are not installed
p_to_install <- p_needed[!(p_needed %in% packages)]

# Install missing packages if any
if (length(p_to_install) > 0) {
  install.packages(p_to_install)
}

# Check and load all packages, report which ones failed
results <- sapply(p_needed, require, character.only = TRUE)

# Display summary of installed and loaded packages
successfully_loaded <- names(results[results == TRUE])
failed_to_load <- names(results[results == FALSE])

cat("Successfully loaded packages:\n")
print(successfully_loaded)

cat("\nPackages failed to load:\n")
print(failed_to_load)

# Set an option for the final document that can be produced from the .Rmd file.
knitr::opts_chunk$set(echo = TRUE)


```

```{css, echo=FALSE}
/* Pink highlight box */
.box-pink {
  background-color: #E72E6B;
  color: white;
  padding: 15px 20px;
  border-radius: 6px;
  margin: 15px 0;
}

/* Blue highlight box */
.box-blue {
  background-color: #6EC7D9;
  color: black;
  padding: 15px 20px;
  border-radius: 6px;
  margin: 15px 0;
}

.box-pink a { color: white; text-decoration: underline; }
.box-blue a { color: #1a1a1a; text-decoration: underline; }
```

::: box-blue
**What You Have Learned So Far**

-   Different harmonisation approaches and their trade-offs
-   How to assess whether survey variables are comparable
-   Which strategies can be used to harmonise response scales
:::

::: box-pink
**What You Will Do in This Section**

In this section you will learn

-   How to download data from the GESIS repository, including I4NG harmonised datasets, directly from within your R workflow using the rgesis package
-   Understand the variable structure: substantive vs. background variables
-   Learn how to explore a variable: frequencies, country coverage, year coverage, and source surveys
-   Understand how harmonisation adds value
-   How to merge datasets across policy pillars using the unique respondent ID
:::

### The Five I4NG Datasets

The Infra4NextGen (I4NG) project provides ready-to-use harmonised datasets. Each dataset corresponds to one of the five thematic pillars of EU youth policy.

Each dataset is identified by a ZA study number. This is a unique ID used by the GESIS Data Archive to help you cite the data and locate it quickly.

| ZA Number | Pillar      | Time Period | Substantive Variable Prefix |
|-----------|-------------|-------------|-----------------------------|
| ZA9070    | **G**reen   | 1982–2023   | `g_`                        |
| ZA9071    | **S**trong  | 1981–2024   | `s_`                        |
| ZA9074    | **H**ealthy | 1981–2023   | `h_`                        |
| ZA9073    | **D**igital | 1981–2023   | `d_`                        |
| ZA9072    | **E**qual   | 1985–2023   | `e_`                        |

In our data page, you will find

-   Data download links

-   Topics that are covered in each pillar

-   Documentation of the datasets

-   Coverage maps per harmonised item and survey

### Understanding Harmonisation Documentation

::: box-pink
**Always Consult the Documentation**

Before analysing harmonised variables, check the GESIS documentation to understand:

-   How variables were harmonised across different source surveys
-   What each variable measures and its conceptual definition
-   Survey-specific notes and caveats about comparability
-   Response scales and coding schemes used

Access documentation at: [I4NG Harmonisation Gateway](https://infra4nextgen.com/harmonisationgateway/data.html)
:::

::: box-pink
**Understanding Variable Naming in I4NG Data**

Variables are organised into two types:

-   **Substantive variables:** These are the variables you are likely most interested in analysing because they carry substantial content. They always start with the pillar letter so you can distinguish them easily.

**Format: Pillar \_ Concept \_ Label**

`e_` = Equal, `d_` = Digital, `g_` = Green, `h_` = Healthy, `s_` = Strong. These are the harmonised survey questions.

-   **Background/protocol variables** are everything else. Background variables describe the respondent and survey context (e.g., `country_I4NG`, `year`, `caseID_I4NG`). Protocol variablesidentify the study (e.g., `studyno_I4NG`, `survey`, `survey_year_title`, `doi_I4NG`).\
:::

## Hands-on

This section will show you how to access and work with I4NG harmonised datasets for your research.\

If you are not an R user: You can still follow along! We have commented on the code chunks. By reading the comments (text starting with \#) and further explanations, you should be able to understand the logic of data processing.

::: box-pink
To access the datasets, you need a free GESIS account. Create one at [here](https://login.gesis.org).
:::

### Setup: Install and Load Packages

**One-time setup**\
You only need to run the installation lines **once**. After that, just load the packages at the start of each R session.

```{r install-packages, eval=FALSE}
# --- ONE-TIME SETUP ---
# Only run these lines if you haven't installed the packages yet
# This chunk takes some time

install.packages("pak")        #  package installer
 pak::pkg_install("jslth/rgesis") # Install rgesis from GitHub

 install.packages(c("haven", "dplyr", "labelled"))
```

```{r load-packages}
# --- LOAD PACKAGES (run every session) ---

library(rgesis)    # Access GESIS data directly from R
library(haven)   # Read SPSS (.sav) files 
library(dplyr)     # Data manipulation 
library(labelled) # Work with variable labels 
library(tidyr)     # pivot_wider for coverage matrices

```

```{r authenticate, eval=FALSE, }
# Opens a browser window for you to log in 
gesis_auth()
```

::: box-pink
-   The `rgesis` package handles authentication automatically via your browser no manual token management needed.
-   When successful, you will see: Successfully performed GESIS login.
:::

### Download the I4NG Equal Dataset (ZA9072)

Now, let us download your first dataset. Remember the ZA numbers from the datasets table? These are ID numbers that uniquely identify datasets in the repository and allow you to download them directly from R. We will use the **Make it Equal dataset (ZA9072)** for our first working example.

```{r download-data-purpose, eval=FALSE}

# Download the SPSS version of the Equal dataset from GESIS 
I4NG_equal_file <- gesis_data("ZA9072", select = "\\.sav")
## IMPORTANT: Run this chunk manually in the console (don't knit it)
#When you run gesis_data(), R will ask you to specify the purpose of your data use: 
#ℹ Please specify a purpose for the use of the research data. 
 
# 1: for final thesis of the study programme (e.g. Bachelor/Master thesis) 
# 2: for research with a commercial mission 
# 3: for non-scientific purposes 
# 4: for further education and qualification 
# 5: for scientific research (incl. doctorate) 
# 6: in the course of my studies 
# 7: in a course as a lecturer 
   
# In the console, type the number that best matches your purpose and press Enter. 

 

```

```{r download-data, eval=FALSE}

I4NG_equal_data <- read_sav(I4NG_equal_file, user_na = TRUE) %>% 
  as.data.frame() 
# user_na = TRUE tells R to treat SPSS missing value codes (-9, -99, -999) as NA.  
# This means  !is.na() will automatically exclude all three types later in your analysis.
 

# How big is it? It is always good practice to check the size of your dataset after loading it. 

# The first number is the number of rows (respondents), and the second is the number of columns (variables).  
dim(I4NG_equal_data) 

# Create folder and save
dir.create(here("data"), showWarnings = FALSE)
saveRDS(I4NG_equal_data, here("data", "I4NG_equal_data.rds"))  # save for knitting


```

```{r load-equal-data}
# LOAD THE SAVED DATA FILE
# This will only work after you've manually run the download-data chunk above
I4NG_equal_data <- readRDS(here("data", "I4NG_equal_data.rds")) #load 
```

::: box-blue
**What just happened?**

-   `gesis_data()` connected to GESIS and downloaded the SPSS file for ZA9072
-   `read_sav()` loaded the file into R as a data frame, preserving all SPSS labels
-   `dim()` tells you the size: **rows = respondents**, **columns = variables**
:::

### Explore the Variable Structure

#### View All Variables and Their Labels

Before doing any analysis, let us explore the contents of the dataset. It helps to understand the data structure and which variables are available.

```{r explore-labels-quick}


# View All Variables and Their Labels 
#  Quick named list of all variable labels 
var_label(I4NG_equal_data) 

```

```{r explore-labels-table}

# Detailed table: variable name + label + type + missing count 
#It creates a summary table showing:  
# - The variable name  
# - The descriptive label (the question asked 
# - The data type (e.g., numeric or factor) 
# - The count of missing values (NA) 

detailed_table <- data.frame( 
   variable  = names(I4NG_equal_data), 
   label     = sapply(I4NG_equal_data, function(x) attr(x, "label")), 
   type      = sapply(I4NG_equal_data, function(x) class(x)[1]), 
   n_missing = sapply(I4NG_equal_data, function(x) sum(is.na(x))) 
) 
  

# View the summary table 

print(detailed_table) 

```

#### Identify Substantive vs. Background Variables

Remember the two variable types in the dataset? Here is where you see how many of each exist in your dataset. Substantive variables are the items you will most likely analyse, and running this code tells you at a glance how much content the pillar covers. Background and protocol variables matter too, as knowing which are available tells you how to contextualise your results and their sources.

```{r variable-types}
# Get all variable names
all_vars <- names(I4NG_equal_data)
print(all_vars)

# Substantive harmonised variables (start with "e_") 
substantive_vars <- all_vars[grepl("^e_", all_vars)] 
print(paste("Number of substantive variables:", length(substantive_vars))) 
print(substantive_vars) 
 
# Background and protocol variables (everything else) 
back_prot_vars <- all_vars[!grepl("^e_", all_vars)] 
print(paste("Number of background/protocol variables:", length(back_prot_vars))) 
head(back_prot_vars, 20) 

```

### The Unique Respondent ID – `caseID_I4NG`

::: box-blue
**The Key to Cross-Pillar Analysis: `caseID_I4NG`**

`caseID_I4NG` is the unique respondent identifier that is **consistent across all five I4NG datasets**. This is the key variable you use to merge data from different policy pillars.
:::

```{r case-id}
# Preview the respondent ID
caseID <- I4NG_equal_data %>%
  select(caseID_I4NG)

head(caseID, 10)


# Check whether IDs are  unique 
I4NG_equal_data %>% 
   summarise( 
     total_rows     = n(), 
     unique_ids     = n_distinct(caseID_I4NG), 
     ids_are_unique = n() == n_distinct(caseID_I4NG)  # should reads as TRUE
   ) 

```

### Explore One Variable in Detail

We will use `e_wr_govn` (Government Responsibility in Income Redistribution) as our example variable throughout this section.

Please open the documentation, where you can see the question text, response scales, harmonisation decisions, and more for this item. We highly suggest keeping the documentation open while working with the datasets.

::: box-pink
**Missing Value Codes in I4NG Data**

I4NG datasets use standardised missing value codes. Always filter these out before any content-related analysis:

-   **`-9`** = Missing values from source datasets
-   **`-99`** = Not applicable for this item\
-   **`-999`** = Inapplicable from source datasets
:::

Start with the overall distribution across all surveys and countries combined.

```{r count}
I4NG_equal_data %>% group_by(e_wr_govn) %>% summarise(n = n())
```

What do you see above? The variable has both valid responses (1–5 scale) and missing value codes: -9, -99, -999. Also note the values 2.333… and 3.666… these are harmonised intermediate values produced by the linear stretch method when a source scale was mapped onto the 1–5 target scale.

```{r frequencies }

# Overall frequency table 
I4NG_equal_data %>% 
  filter(!is.na(e_wr_govn)) %>%  
   group_by(e_wr_govn) %>% 
  summarise(n = n()) %>%  #Count how many people chose each answer (e.g., how many said '1'). 
  mutate(pct = round(n / sum(n) * 100, 1)) %>%  # A new column ('pct') to show the relative share. Round to 1 decimal place (e.g., 25.4%) for readability. 
  arrange(e_wr_govn)   # Sort: Organize the table by the answer code 

 
```

You can filter missing codes if you are merely interested in the content-related variables. Because we used user_na = TRUE, !is.na() handles all three of missing values (-9, -99, -999) at once.

```{r coverage}
#Which Source Surveys Cover This Variable? 

#The harmonised data pools from multiple original surveys. 

# How many source surveys and respondents in total? 
I4NG_equal_data %>% 
   summarise( 
     n_survey_year_titles = n_distinct(survey_year_title), 
     n_surveys            = n_distinct(survey), 
     n_years              = n_distinct(year), 
     total_respondents    = n() 
   )
```

```{r valid-responses}
#  Which surveys have valid data for e_wr_govn, and how many respondents?
I4NG_equal_data %>% 
  filter(!is.na(e_wr_govn)) %>% 
  group_by(survey_year_title) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

```

```{r frequencies-by-survey}
##Frequencies Split by Survey 

I4NG_equal_data %>% 
  filter(!is.na(e_wr_govn)) %>% 
  group_by(survey_year_title, e_wr_govn) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  group_by(survey_year_title) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>% 
  arrange(survey_year_title, e_wr_govn)
```

```{r country coverage}
#Country Coverage for This Variable 

# How many unique countries have valid data for e_wr_govn? 
I4NG_equal_data %>% 
   filter(!is.na(e_wr_govn)) %>% 
   summarise(n_countries = n_distinct(country_I4NG))
```

```{r n-per-country}
# N per country 
I4NG_equal_data %>% 
  filter(!is.na(e_wr_govn)) %>% 
  group_by(country_I4NG) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n)) 
```

```{r countries}
# How many unique countries per source survey? 
I4NG_equal_data %>% 
   filter(!is.na(e_wr_govn)) %>% 
   group_by(survey_year_title) %>% 
   summarise(n_countries = n_distinct(country_I4NG)) %>% 
   arrange(desc(n_countries))
```

```{r country-year}
#Country × Year Coverage Matrix 

#This shows which countries are available in which years. 

I4NG_equal_data %>% 
  filter(!is.na(e_wr_govn)) %>% 
  group_by(country_I4NG, year) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  tidyr::pivot_wider( 
    names_from  = year, 
    values_from = n, 
    values_fill = 0 
  )

```

::: box-blue
**What to look for in these outputs?**

Are there surveys that only cover certain years? This is normal, different programmes run at various times.

Are some cells -999? That means no respondents from that country answered this item in that year.
:::

#### Coverage Across Surveys, Countries and Time

So far, we have been working with the Equal dataset and one variable. Now, examine what the Green dataset looks like, specifically how many surveys, countries, and years contribute to a given variable, and how that compares to using any single source survey on its own.

We use `g_ep_clai` (“Many claims about the environment are exaggerated”) for our example.

```{r download-green-data, eval=FALSE}
# IMPORTANT: Run this chunk manually in the console (don't knit it)
I4NG_green_file <- gesis_data("ZA9070", select = "\\.sav") 
I4NG_green_data <- read_sav(I4NG_green_file, user_na = TRUE) %>% 
  as.data.frame()




# Create folder and save
dir.create(here("data"), showWarnings = FALSE)
saveRDS(I4NG_green_data, here("data", "I4NG_green_data.rds"))  # save for knitting





```

Now, let us find which single survey has the most respondents for this variable. By 'best,' we mean the survey with the largest sample size (highest N) for `g_ep_clai`. We will then compare this single survey's coverage against the combined harmonised dataset to show you the real power of pooling data across multiple surveys.

```{r load-green-data}
# LOAD THE SAVED DATA FILE
# This will only work after you've manually run the download-data chunk above
I4NG_green_data <- readRDS(here("data", "I4NG_green_data.rds")) #load
```

```{r  single-best-survey}
# --- The single best survey --- 
I4NG_green_data %>% 
  filter(!is.na(g_ep_clai)) %>% 
  group_by(survey_year_title) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n)) %>% 
  head(1)
```

```{r country-survey}
# --- Countries per survey --- 
I4NG_green_data %>% 
   filter(!is.na(g_ep_clai)) %>% 
   group_by(survey_year_title) %>% 
   summarise(n_countries = n_distinct(country_I4NG)) %>% 
   arrange(desc(n_countries))
```

```{r country-coverage}
# Total country coverage 
I4NG_green_data %>% 
   filter(!is.na(g_ep_clai)) %>% 
   summarise(n_countries = n_distinct(country_I4NG))
```

```{r total}
# --- The harmonised total --- 
I4NG_green_data %>% 
   filter(!is.na(g_ep_clai)) %>% 
   summarise( 
     total_n     = n(), 
     n_countries = n_distinct(country_I4NG), 
     n_years     = n_distinct(year), 
     n_surveys   = n_distinct(survey_year_title) 
   )
```

::: box-blue
**What to look for?**

Compare the largest single survey’s N and country count against the harmonised totals. You will typically find that no single source survey covers as many countries, years, or respondents as the combined dataset. This is the practical consequence of pooling across programmes by harmonisation: broader coverage, at the cost of having to account for source heterogeneity in your analysis.

What happens when you want to analyse two variables from different policy pillars? This is where the `caseID_I4NG` becomes crucial. Let us see how to merge datasets and what trade-offs you will encounter.
:::

### Merging Datasets Across Pillars

```{r frequency-g}
#Explore a Variable from the Green Dataset 

#Before merging, let us explore g_ep_clai the same way we did for e_wr_govn. 

# Overall frequencies for g_ep_clai  
I4NG_green_data %>% 
   filter(!is.na(g_ep_clai)) %>%
  group_by(g_ep_clai) %>% 
  summarise(n = n()) %>%
   mutate(pct = round(n / sum(n) * 100, 1)) %>% 
   arrange(g_ep_clai)
```

```{r freq-by-survey}
# Frequencies split by survey 
I4NG_green_data %>% 
   filter(!is.na(g_ep_clai)) %>% 
   group_by(survey_year_title, g_ep_clai) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  group_by(survey_year_title) %>%
   mutate(pct = round(n / sum(n) * 100, 1)) %>% 
   arrange(survey_year_title, g_ep_clai)

```

Who Answered Both Items `g_ep_clai` and `e_wr_govn`?

The code below identifies respondents who provided valid responses to both items. It selects the ID and variable from each dataset, joins them on the shared `caseID_I4NG` identifier, and drops anyone with a missing value for either item.

```{r both}
#  Find caseIDs with valid responses to BOTH variables 
valid_ids <- I4NG_equal_data %>% 
   select(caseID_I4NG, e_wr_govn) %>% 
   inner_join( 
     I4NG_green_data %>% select(caseID_I4NG, g_ep_clai), 
     by = "caseID_I4NG" 
   ) %>% 
   filter(!is.na(e_wr_govn), !is.na(g_ep_clai)) %>% 
   pull(caseID_I4NG) 
 
# How many respondents answered BOTH? 
length(valid_ids)
```

::: box-blue
**What you should see?**

The number of respondents who answered both items is substantially smaller than the total number of each of the individual datasets. With the code below, checking which surveys those respondents come from confirms this directly.
:::

```{r}
# Which surveys do these respondents come from? 
I4NG_equal_data %>% 
  filter(caseID_I4NG %in% valid_ids) %>% 
  group_by(survey_year_title) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n))
 


```

```{r}
# How many unique surveys contributed to the overlap? 
I4NG_equal_data %>% 
   filter(caseID_I4NG %in% valid_ids) %>% 
   summarise(unique_surveys = n_distinct(survey_year_title))
```

::: box-blue
This is the core trade-off: individual-level analysis is possible but comes at the cost of a much smaller sample, limited to surveys that include both items.
:::

Now, let us put it all together. You have explored individual variables and understood the trade-offs in coverage. Here is how to merge data from different pillars using the `caseID_I4NG` identifier. You can merge these datasets because the `caseID_I4NG` identifier ensures that you are matching the same individuals across different datasets.

### Append a Variable from the Green into the Equal Dataset

Once you have decided on your analysis, you can merge the `g_ep_clai` variable from the Green dataset into the Equal dataset using `caseID_I4NG`. All Equal rows are retained; matching respondents get the `g_ep_clai` value, others get NA.

```{r}

# Append g_ep_clai from Green into the Equal dataset 
I4NG_equal_data <- I4NG_equal_data %>% 
   left_join( 
     I4NG_green_data %>% select(caseID_I4NG, g_ep_clai), 
     by = "caseID_I4NG" 
   ) 
 
# Verify: g_ep_clai is now in the Equal dataset 
I4NG_equal_data %>% 
   filter(!is.na(e_wr_govn), !is.na(g_ep_clai)) %>% 
   select(caseID_I4NG, e_wr_govn, g_ep_clai) %>% 
   head(10) 
 
# Check: how many rows now have valid values for both? 
I4NG_equal_data %>% 
   filter(!is.na(e_wr_govn), !is.na(g_ep_clai)) %>% 
   nrow() 

  
```

### Take-home Exercise

Try the following steps on your own to consolidate what you have learned:Use the Strong (ZA9071) and the Digital datasets (ZA9073).

-   Download both datasets using gesis_data()
-   Identify substantive harmonised variables.
-   Pick one variable from each dataset. Consult the documentation before starting the analysis. Check: overall frequency, source survey coverage, country coverage, year coverage.
-   Find how many respondents answered both of your chosen variables. Which specific surveys (e.g., ESS 2016, Eurobarometer 2020) contributed to this overlap?
-   Append your chosen Strong variable to the Digital dataset so you can analyse them together in one data frame

```{r , eval=FALSE}
# --- Step 1: Download both datasets --- 
I4NG_strong_file <- gesis_data("ZA9071", select = "\\.sav") 

I4NG_strong_data <- read_sav(I4NG_strong_file, user_na = TRUE) %>%
as.data.frame()

I4NG_digital_file <-  

I4NG_digital_data <-
```

```{r , eval=FALSE}
# --- Step 2: Identify substantive harmonised variables --- 

#  Hint: Strong variables start with "s_", Digital with "d_"
```

```{r , eval=FALSE}
#  --- Step 3: Pick one variable from each dataset and explore it --- 

#  Replace s_your_variable and d_your_variable with your chosen variables 

#  Consult the documentation before starting: https://infra4nextgen.com/harmonisationgateway/data.html 

 

#  Overall frequency 

 

#  Source survey coverage 

 

#  Country coverage 

 

#  Year coverage
```

```{r , eval=FALSE}
#  --- Step 4: Find respondents who answered BOTH variables --- 

#  Hint: use inner_join() on caseID_I4NG, then filter out NAs on both variables
```

```{r , eval=FALSE}
#  --- Step 5: Append your Strong variable to the Digital dataset --- 

#  Hint: use left_join() on caseID_I4NG

 
 
 

  


```

### Summary

::: box-blue
**You now know how to:**

-   Authenticate with GESIS and download I4NG harmonised datasets
-   Distinguish substantive from background and protocol variables
-   Explore a variable: frequencies overall, by survey, by country, by year
-   Understand which source surveys contribute to each variable
-   Use caseID_I4NG to merge data across policy pillars
:::

**Next:** Go to **next section in the course** to learn how to assess the quality of harmonised measures and understand the consequences of choosing certain harmonisation strategies for the data.

------------------------------------------------------------------------

*For questions or support, contact the I4NG team at [harmonisation.i4ng\@gesis.org](mailto:harmonisation.i4ng@gesis.org)*
